Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-09-26 doc/invt.tex
add catxerr.py to combine xerr.dat from different directories

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors, which can be
reduced by gradually decreasing the updating magnitude.\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, morozov2009, zhou2008, min2007,
  poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}

@@

We can split the $f_i(t)$ into
a deterministic averaged part, $\langle f_i(t) \rangle$,
and a random fluctuating part, $\xi_i(t)$,
%
\begin{equation}
  f_i(t) =
  \langle f_i(t) \rangle
  +
  \xi_i(t)
  .
  \notag
  %\label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consisting of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For a sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
[cf. Eq. \eqref{eq:pi_p_V}]
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle f_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { \rho_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { \rho_i }
  =
  \frac{                          e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n \rho_j \, e^{-v_{*j}(t)} }
  %\notag
  %\\
  %&
  \approx
  1 - v_{*i}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion,
and used Eq. \eqref{eq:x_def} in the last step.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { \rho_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n \rho_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n \rho_j \, v_{*j} }
.
$$
But
$$
\sum_{ j = 1 }^n \rho_j \, v_{*j}
=
\sum_{ j = 1 }^n \rho_j \, v_j - \sum_{j=1}^n \bav{v}  = 0.
$$
}%
%
Deducting the bin average
and noticing that
$\bav{f}(t) = \sum_{i=1}^n \rho_i [h_i(t)/\rho_i] = \sum_{i=1}^n \delta_{i, i(t)} = 1$,
we get
$\langle f_{*i}(t) \rangle = \langle f_i(t) \rangle - 1 \approx - v_{*i}(t)$,
%
%\begin{align*}
%  \langle f_{*i}(t) \rangle
%  \equiv
%  \langle f_i(t) \rangle -
%  \bav{f}(t)
%  =
%  \langle f_i(t) \rangle - 1
%  \approx
%  - v_{*i}(t)
%  ,
%\end{align*}
%
and
%
\begin{equation}
  f_{*i}(t) \approx - v_{*i}(t) + \xi_{*i}(t)
  .
  \label{eq:sh_ave}
\end{equation}

@@

From Eqs.
\eqref{eq:vt_diffeq} and \eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  %=
  %\alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \xi_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \xi_{*i}(t) \right]
  .
  %\notag
  \label{eq:dxdt_singlebin}
\end{equation}
%
Without the random part, $\xi_{*i}(t)$,
the deviation of the bias potential, $v_{*i}(t)$,
would tend to $0$ at long times, reaching convergence.
%
The formal solution of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
  v_{*i}(T)
  =
  v_{*i}(0) \, e^{-q(T)}
  +
  \int_0^T
    \dot{\theta}\bigl( q(t) \bigr) \, \xi_{*i}(t) \, dt,
  \label{eq:vi_solution}
\end{equation}
%
where
%
$q(t) \equiv \int_0^t \alpha(t') \, dt'$,
%
and
%
\begin{align}
  \theta(q')
  &\equiv
  e^{q' - q(T)}.
  \label{eq:theta_def}
\end{align}

@@

From Eqs. \eqref{eq:error_def} and
\eqref{eq:vi_solution},
we can write the total error at the end of period $T$
as a sum of a residual error, $E_R(T)$ (from the decay of the initial error)
and an asymptotic error, $E_A(T)$:
%
\begin{align}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T)
  ,
  \label{eq:error_tot}
\end{align}
%
where
\begin{align}
  \Err_R(T)
  &= \Err(0) \, e^{-2 \, q(T)}
  ,
  \label{eq:ER_sbin}
  \\
  \Err_A(T)
  &=
  \sum_{i=1}^n \rho_i
    \int_0^T\!\!\!\int_0^T
    \!\!
    \langle
      \xi_{*i}(t) \xi_{*i}(t')
    \rangle
    \dot \theta\bigl( q(t) \bigr)
    \dot \theta\bigl( q(t') \bigr)
    \, dt dt'
  \notag \\
  &= \Gamma \, \int_0^T \dot \theta^2\bigl( q(t) \bigr) \, dt
  .
  \label{eq:EA_sbin}
\end{align}
In Eq. \eqref{eq:EA_sbin},
we have used Eq. \eqref{eq:Gi_def} and introduced
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
\approx \sum_{i=1}^n \rho_i \int_{-\infty}^{\infty} \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt$.
%\begin{equation}
%  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
%  =
%  \sum_{i=1}^n \rho_i \int_0^T
%    \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt
%  .
%  \notag%\label{eq:Gamma_from_G}
%\end{equation}

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme,
Eq. \eqref{eq:mbin_update} [using $u_i(t)$ instead of $v_i(t)$],
as
%
\begin{equation}
  {\tilde v}_k(t + 1) =
  {\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
  {\tilde f}_k(t)
  ,
  \label{eq:vkupdate}
\end{equation}
%
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \xi}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where
we have used Eq. \eqref{eq:xtstar} and
the transformed version of Eq. \eqref{eq:sh_ave}
in the second step.

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error. % in Sec. \ref{sec:optinitalpha}.
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \theta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \theta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  %\notag
  \label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Here, the Hamiltonian coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{goldstein, *landau_mechanics, *arnold, *jose}
which means that the asymptotic error grows
linearly with time at a rate of $\mathcal L$.
We may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
Equation \eqref{eq:Lagrangian_const} serves
as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, Eq. \eqref{eq:Lagrangian_const}
gives an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
This solution can be characterized geometrically,
as shown in Appendix \ref{sec:schedule_geometry}.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^{\!\!2}
  .
\label{eq:error_asym2}
\end{equation}

@@

%
From Eqs. \eqref{eq:error_res},
\eqref{eq:xt2_eql},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ that minimizes $\Err_R(T) + \Err_A(T)$
satisfies
%
\begin{align}
  \Delta\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &\quad
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
  \label{eq:opt_qT}
\end{align}
%where
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{We can write
\begin{align*}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  &=
  \frac { a_0 \, \Gamma_k \, \lambda_k } { 2 }
  +
  \Delta_k
  e^{-2\, \lambda_k a_0 T_0}
  ,
\end{align*}
%
where $\Delta_k \equiv \tilde u_k^2 - \frac{a_0 \lambda_k \Gamma_k}{2}$.
%
Let $q_T \equiv q(T)$,
\begin{align*}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -2\sum_{k=1}^{n-1} \lambda_k
  \left\langle
    \tilde v_k^2(0)
  \right\rangle \,
  e^{-2 \, \lambda_k \, q_T}
  \\
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \Delta_k e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{align*}
Then solve $\partial (\Err_R + \Err_A) / \partial q_T = 0$,
\begin{align*}
  \frac{ M\bigl( q_T \bigr) } { T } C_M
  -\frac{a_0}{2} \, M^2\bigl( q_T \bigr)
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \Delta_k \, e^{-2\lambda_k \, (q_T + a_0 T_0)}
  =0
  .
\end{align*}
}
%
We can rewrite the condition in terms of
the initial updating magnitude
by Eq. \eqref{eq:Lagrangian_const}, %\eqref{eq:mQ_invTa},
\begin{align}
  \alpha(t = 0)
  %=
  %\frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  \stackrel{\epsilon_k \to 0}{=\joinrel=\joinrel=}
  \frac{ a_0 }
       { 2 }
  .
  %\notag
  \label{eq:half_alpha0}
\end{align}
%
%
The equality in the last step is achievable
if the initial bias, $\epsilon_k$, is negligible,
%which is often true if no eigenvalue is close to $0$,
and in this limit, the optimal initial updating magnitude
is half of the value
in the preliminary run.\note{This factor, $1/2$,
  happens to coincide with the
  recommended reduction factor of the updating magnitude
  at stage transitions
  in the original WL algorithm\cite{
  wang2001, wang2001pre}.  }%
%[cf. Eq. \eqref{eq:t0_sbin}].

@@

A commonly-used multiple-bin updating scheme in metadynamics
is the Gaussian updating scheme,
whose updating matrix can be characterized by
a rigid Gaussian function centered on the current bin
(with some boundary adjustment).
%to satisfy the constraints discussed in Sec. \ref{sec:updating-matrix},
%see Fig. \ref{fig:mat}(a) for an example.
%
In terms of the continuous collective variable, $z$,
the eigenvalues of this scheme are given by\cite{bussi2006}
\begin{align}
  \lambda_k
  =
  \exp\left(
        -
        \frac{ \pi^2 \, \sigma_z^2 \, k^2 }
             { 2 \, \Delta z^2 }
      \right)
  ,
  %\notag
  \label{eq:lambda_Gaussian}
\end{align}
where $\sigma_z$ is the width of the Gaussian
and $\Delta z = n \, \delta z \, g/2$
is the effective span
with $g = 1$ for a periodic variable or $2$ otherwise
(cf. Appendix \ref{sec:Gaussian_math}).

>>>>

While adaptive FDS simulations
can often sample the desired distribution,
the frequent updates cause the bias potential to fluctuate constantly,
and disrupt the underlying equilibrium sampling.
%
To reduce the sampling error,
the updating magnitude is often decreased gradually.\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, morozov2009, zhou2008, min2007,
  poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}

@@

We can split the $f_{*i}(t)$ into
a deterministic averaged part, $\langle f_{*i}(t) \rangle$,
and a random fluctuating part, $\xi_{*i}(t)$,
%
\begin{equation}
  f_{*i}(t) =
  \langle f_{*i}(t) \rangle
  +
  \xi_{*i}(t)
  .
  \notag
  %\label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consisting of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For a sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
[cf. Eq. \eqref{eq:pi_p_V}]
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle f_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { \rho_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { \rho_i }
  =
  \frac{                          e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n \rho_j \, e^{-v_{*j}(t)} }
  %\notag
  %\\
  %&
  \approx
  1 - v_{*i}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion,
and used Eq. \eqref{eq:x_def} in the last step.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { \rho_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n \rho_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n \rho_j \, v_{*j} }
.
$$
But
$$
\sum_{ j = 1 }^n \rho_j \, v_{*j}
=
\sum_{ j = 1 }^n \rho_j \, v_j - \sum_{j=1}^n \bav{v}  = 0.
$$
}%
%
Deducting the bin average
and noticing that
$\bav{f}(t) = \sum_{i=1}^n \rho_i [h_i(t)/\rho_i] = \sum_{i=1}^n \delta_{i, i(t)} = 1$,
we get
$\langle f_{*i}(t) \rangle = \langle f_i(t) \rangle - 1 \approx - v_{*i}(t)$,
%
%\begin{align*}
%  \langle f_{*i}(t) \rangle
%  \equiv
%  \langle f_i(t) \rangle -
%  \bav{f}(t)
%  =
%  \langle f_i(t) \rangle - 1
%  \approx
%  - v_{*i}(t)
%  ,
%\end{align*}
%
and
%
\begin{equation}
  f_{*i}(t) \approx - v_{*i}(t) + \xi_{*i}(t)
  .
  \label{eq:sh_ave}
\end{equation}

@@

From Eqs.
\eqref{eq:vt_diffeq} and \eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  %=
  %\alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \xi_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \xi_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is % of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
  v_{*i}(T)
  =
  v_{*i}(0) \, e^{-q(T)}
  +
  \int_0^T
    \dot{\theta}\bigl( q(t) \bigr) \, \xi_{*i}(t) \, dt,
  \label{eq:vi_solution}
\end{equation}
%
where
%
$q(t) \equiv \int_0^t \alpha(t') \, dt'$,
%
and
%
\begin{align}
  \theta(q')
  &\equiv
  e^{q' - q(T)}.
  \label{eq:theta_def}
\end{align}
%
Thus, without the random part, $\xi_{*i}(t)$,
the deviation of the bias potential, $v_{*i}(t)$,
would tend to $0$ at long times, reaching convergence,
if $\alpha(t)$ decays sufficiently slowly,
i.e. $q(T) \to \infty$ as $T \to \infty$.

@@

From Eqs. \eqref{eq:error_def} and
\eqref{eq:vi_solution},
we can write the total error at the end of period $T$
as a sum of a residual error, $E_R(T)$ (from the decay of the initial error)
and an asymptotic error, $E_A(T)$:
%
\begin{align}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T)
  ,
  \label{eq:error_tot}
\end{align}
%
where
\begin{align}
  \Err_R(T)
  &= \Err(0) \, e^{-2 \, q(T)}
  ,
  \label{eq:ER_sbin}
  \\
  \Err_A(T)
  &=
  \sum_{i=1}^n \rho_i
    \int_0^T\!\!\!\int_0^T
    \!\!
    \langle
      \xi_{*i}(t) \xi_{*i}(t')
    \rangle
    \dot \theta\bigl( q(t) \bigr)
    \dot \theta\bigl( q(t') \bigr)
    \, dt dt'
  \notag \\
  &= \Gamma \, \int_0^T \dot \theta^2\bigl( q(t) \bigr) \, dt
  .
  \label{eq:EA_sbin}
\end{align}
In Eq. \eqref{eq:EA_sbin},
we have used Eq. \eqref{eq:Gi_def} and introduced
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
%$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i
%\approx \sum_{i=1}^n \rho_i \int_{-\infty}^{\infty} \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt$.
\begin{equation}
  \Gamma
  \equiv
  \sum_{i=1}^n \rho_i \, G_i
  \approx
  \sum_{i=1}^n \rho_i \int_0^T
    \langle \xi_{*i}(t) \xi_{*i}(0) \rangle \, dt
  .
  \notag%\label{eq:Gamma_from_G}
\end{equation}

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme,
Eq. \eqref{eq:mbin_update} [using $u_i(t)$ instead of $v_i(t)$],
as
%
\begin{equation}
  {\tilde v}_k(t + 1) =
  {\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
  {\tilde f}_k(t)
  .
  \label{eq:vkupdate}
\end{equation}
%
In continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \xi}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where
we have Fourier transformed Eq. \eqref{eq:sh_ave}
and used Eq. \eqref{eq:xtstar}
in the second step.

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error. % in Sec. \ref{sec:optinitalpha}.
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \theta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \theta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  %\notag
  \label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Here, the Hamiltonian coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{goldstein, *landau_mechanics, *arnold, *jose}
i.e. the asymptotic error grows
linearly with time at a rate of
$\mathcal L = \mathcal H$.
%
We may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
Equation \eqref{eq:Lagrangian_const} serves
as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, Eq. \eqref{eq:Lagrangian_const}
gives an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
This solution can be characterized geometrically,
as shown in Appendix \ref{sec:schedule_geometry}.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^{\!\!2}
  .
\label{eq:error_asym2}
\end{equation}

@@

%
From Eqs. \eqref{eq:error_res},
\eqref{eq:xt2_eql},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ that minimizes $\Err_R(T) + \Err_A(T)$
satisfies
%
\begin{align}
  \Delta\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} \!\!\! M(\bar q) \, d\bar q
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &\quad
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
  \label{eq:opt_qT}
\end{align}
%where
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{We can write
\begin{align*}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  &=
  \frac { a_0 \, \Gamma_k \, \lambda_k } { 2 }
  +
  \Delta_k
  e^{-2\, \lambda_k a_0 T_0}
  ,
\end{align*}
%
where $\Delta_k \equiv \tilde u_k^2 - \frac{a_0 \lambda_k \Gamma_k}{2}$.
%
Let $q_T \equiv q(T)$,
\begin{align*}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -2\sum_{k=1}^{n-1} \lambda_k
  \left\langle
    \tilde v_k^2(0)
  \right\rangle \,
  e^{-2 \, \lambda_k \, q_T}
  \\
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \Delta_k e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{align*}
Then solve $\partial (\Err_R + \Err_A) / \partial q_T = 0$,
\begin{align*}
  \frac{ M\bigl( q_T \bigr) } { T } C_M
  -\frac{a_0}{2} \, M^2\bigl( q_T \bigr)
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \Delta_k \, e^{-2\lambda_k \, (q_T + a_0 T_0)}
  =0
  .
\end{align*}
}
%
We can rewrite the condition in terms of
the initial updating magnitude
by Eqs. \eqref{eq:Lagrangian_const} and \eqref{eq:mint}, %\eqref{eq:mQ_invTa},
\begin{align}
  \alpha(t = 0)
  %=
  %\frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{
    M^2\bigl( q(T) \bigr)
  }
  \stackrel{\epsilon_k \to 0}{=\joinrel=\joinrel=}
  \frac{ a_0 }
       { 2 }
  .
  %\notag
  \label{eq:half_alpha0}
\end{align}
%
%
The equality in the last step is achievable
if the initial bias, $\epsilon_k$, is negligible,
%which is often true if no eigenvalue is close to $0$,
and in this limit, the optimal initial updating magnitude
is half of the value
in the preliminary run.\note{This factor, $1/2$,
  happens to coincide with the
  recommended reduction factor of the updating magnitude
  at stage transitions
  in the original WL algorithm\cite{
  wang2001, wang2001pre}.  }%
%[cf. Eq. \eqref{eq:t0_sbin}].

@@

A commonly-used multiple-bin updating scheme in metadynamics
is the Gaussian updating scheme,
whose updating matrix can be characterized by
a rigid Gaussian function centered on the current bin
(with some boundary adjustment).
%to satisfy the constraints discussed in Sec. \ref{sec:updating-matrix},
%see Fig. \ref{fig:mat}(a) for an example.
%
In terms of the continuous collective variable, $z$,
the eigenvalues of the updating matrix are given by\cite{bussi2006}
\begin{align}
  \lambda_k
  =
  \exp\left(
        -
        \frac{ \pi^2 \, \sigma_z^2 \, k^2 }
             { 2 \, \Delta z^2 }
      \right)
  ,
  %\notag
  \label{eq:lambda_Gaussian}
\end{align}
where $\sigma_z$ is the width of the Gaussian
and $\Delta z = n \, \delta z \, g/2$
is the effective span
with $g = 1$ for a periodic variable or $2$ otherwise
(cf. Appendix \ref{sec:Gaussian_math}).
